### 2024-05-29 getting-started.qmd
20240526 - placeholder
20240527 - formatting
20240528 - better data example

# Getting Started with R for Data Analysis {#sec-gettingStarted}

@@

You can install R packages using the following syntax:

@@

@@

```{r}
names(nfl_players)
```

@@

nfl_players$position_group == "RB"
```

@@

nfl_players$position_group != "RB"
```

@@

nfl_players$weight > 300
```

@@

nfl_players$weight < 300
```

@@

nfl_players$weight >= 300
```

@@

nfl_players$weight <= 300
```

@@

nfl_players$position_group %in% c("QB","RB","WR")
```

@@

!(nfl_players$position_group %in% c("QB","RB","WR"))
```

@@

is.na(nfl_players$college_name)
```

@@

!is.na(nfl_players$college_name)
```

@@

nfl_players$position_group == "RB" & nfl_players$weight > 230
```

@@

nfl_players$position_group == "RB" | nfl_players$weight > 230
```

@@

nfl_players$display_name
```

@@

```{r}
nfl_players[,"display_name"]
```

@@

nfl_players$display_name[which(nfl_players$position_group == "RB")]
```

@@

```{r}
nfl_players[which(nfl_players$position_group == "RB"), "display_name"]
```

@@

subsetVars <- c("status","display_name","college_name")

nfl_players[,c(1,2,9)]
nfl_players[,c("status","display_name","college_name")]
nfl_players[,subsetVars]
```

@@

dropVars <- c("status","college_name")

nfl_players[,-c(1,9)]
nfl_players[,!(names(nfl_players) %in% c("status","college_name"))]
nfl_players[,!(names(nfl_players) %in% dropVars)]
```

@@

nfl_players %>%
  select(status, display_name, college_name)

nfl_players %>%
  select(status:college_name)

nfl_players %>%
  select(all_of(subsetVars))
```

@@

nfl_players %>%
  select(-status, -college_name)

nfl_players %>%
  select(-c(status:college_name))

nfl_players %>%
  select(-all_of(dropVars))
```

@@

subsetRows <- c(1,3,5)

nfl_players[c(1,3,5),]
nfl_players[subsetRows,]
nfl_players[which(nfl_players$position_group == "RB"),]
```

@@

nfl_players %>%
  filter(position_group == "RB")

nfl_players %>%
  filter(position_group == "RB", weight <= 250)

nfl_players %>%
  filter(position_group == "RB" | weight >= 250)
```

@@

nfl_players[c(1,3,5), c(1,2,3)]
nfl_players[subsetRows, subsetVars]
nfl_players[which(nfl_players$position_group == "RB"), subsetVars]
```

@@

nfl_players %>%
  filter(position_group == "RB") %>%
  select(all_of(subsetVars))
```

@@

View(nfl_players)
```

@@

```{r}
nfl_players$newVar <- NA
```

@@

```{r}
nfl_players$newVar <- 1:nrow(nfl_players)
```

@@

nfl_players$newVar <- NULL
```

```{r}
mydata <- data.frame(
  ID = c(1:5, 1047:1051),
  cat = sample(
    0:1,
    10,
    replace = TRUE)
)

mydata
```

@@

```{r}
mydata$oldVar1 <- NA
mydata$oldVar1[which(mydata$cat == 0)] <- "dog"
mydata$oldVar1[which(mydata$cat == 1)] <- "cat"

mydata$oldVar2 <- NA
mydata$oldVar2[which(mydata$cat == 0)] <- "no"
mydata$oldVar2[which(mydata$cat == 1)] <- "yes"
```

@@

```{r}
mydata %>%
  mutate(across(c(
    oldVar1:oldVar2),
    ~ case_match(
      .,
      c("dog","no") ~ 0,
      c("cat","yes") ~ 1)))
```

@@

```{r}
mydata <- mydata %>% 
  rename(
    newVar1 = oldVar1,
    newVar2 = oldVar2)
```

@@

mydata <- mydata %>% 
  rename_with(~ varNamesTo, all_of(varNamesFrom))
```

@@

```{r}
mydata$factorVar <- factor(mydata$ID)
mydata$numericVar <- as.numeric(mydata$cat)
mydata$integerVar <- as.integer(mydata$cat)
mydata$characterVar <- as.character(mydata$newVar1)
```

@@

```{r}
mydata %>%
  mutate(across(c(
    ID,
    cat),
    as.numeric))

mydata %>%
  mutate(across(
    ID:cat,
    as.numeric))

mydata %>%
  mutate(across(where(is.factor), as.character))
```

@@

Here are the data in the `nfl_actualStats_offense_weekly` object:

```{r}
dim(nfl_actualStats_offense_weekly)

dim(distinct(nfl_actualStats_offense_weekly, player_id, season, week, .keep_all = TRUE))
```

The data are structured in player-season-week form.
That is, every row in the dataset is uniquely identified by the variables, `player_id`, `season`, and `week`.

Here are the data in the `nfl_expectedFantasyPoints_weekly` object:

```{r}
dim(nfl_players)

dim(distinct(nfl_players, gsis_id, .keep_all = TRUE))
```

@@

fullJoinData <- full_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

leftJoinData <- left_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

rightJoinData <- right_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

innerJoinData <- inner_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

semiJoinData <- semi_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

antiJoinData <- anti_join(
  nfl_actualStats_offense_weekly,
  nfl_players,
  by = c("player_id" = "gsis_id"))

@@

crossJoinData <- cross_join(
  unique(nfl_actualStats_offense_weekly[,c("player_id","player_display_name")]),
  data.frame(season = 2020:2024))

crossJoinData
dim(crossJoinData)
```

@@

@@

dataLong <- nfl_actualStats_offense_weekly %>% 
  select(player_id, player_display_name, season, week, fantasy_points)

@@

Data widened by two variable (`season` and `week`), using `tidyverse`:

dataWide <- dataLong %>% 
  pivot_wider(
    names_from = c(season, week),
    names_glue = "{.value}_{season}_week{week}",
    values_from = fantasy_points)

@@

dataWide <- nfl_actualStats_offense_weekly %>% 
  select(player_id, player_display_name, season, week, recent_team, opponent_team)

@@

dataLong <- dataWide %>% 
  pivot_longer(
    cols = c(recent_team, opponent_team),
    names_to = "role",
    values_to = "team")

@@

>>>>

# Getting Started with `R` for Data Analysis {#sec-gettingStarted}

@@

You can install `R` packages using the following syntax:

@@

```{r}
players <- data.frame(
  ID = 1:10,
  name = c(
    "Ken Cussion",
    "Ben Sacked",
    "Justin Time",
    "Lionel Scrimmage",
    "Drew Blood",
    "Hugo Long",
    "Chase Emdown",
    "Name8",
    "Name9",
    "Name10"),
  position = c("QB","QB","RB","RB","WR","WR","WR","WR","TE","TE"),
  age = c(40, 30, 20, 18, 23, 27, 32, 26, 23, NA)
  )

fantasyPoints <- data.frame(
  ID = c(2, 6, 11, 12),
  fantasyPoints = c(250, 170, 65, 15)
)
```

@@

```{r}
names(nfl_players)

names(players)

names(fantasyPoints)
```

@@

```{r}
players$position == "RB"
```

@@

```{r}
players$position != "RB"
```

@@

```{r}
players$age > 30
```

@@

```{r}
players$age < 30
```

@@

```{r}
players$age >= 30
```

@@

```{r}
players$age <= 30
```

@@

```{r}
players$position %in% c("RB","WR")
```

@@

```{r}
!(players$position %in% c("RB","WR"))
```

@@

```{r}
is.na(players$age)
```

@@

```{r}
!is.na(players$age)
```

@@

```{r}
players$position == "WR" & players$age > 26
```

@@

```{r}
players$position == "WR" | players$age > 23
```

@@

```{r}
players$name
```

@@

```{r}
players[,"name"]
```

@@

```{r}
players$name[which(players$position == "RB")]
```

@@

```{r}
players[which(players$position == "RB"), "name"]
```

@@

```{r}
subsetVars <- c("name","age")

players[,c(2,3)]
players[,c("name","age")]
players[,subsetVars]
```

@@

```{r}
dropVars <- c("name","age")

players[,-c(2,3)]
players[,!(names(players) %in% c("name","age"))]
players[,!(names(players) %in% dropVars)]
```

@@

```{r}
players %>%
  select(name, age)

players %>%
  select(name:age)

players %>%
  select(all_of(subsetVars))
```

@@

```{r}
players %>%
  select(-name, -age)

players %>%
  select(-c(position:age))

players %>%
  select(-all_of(dropVars))
```

@@

```{r}
subsetRows <- c(3,4)

players[c(3,4),]
players[subsetRows,]
players[which(players$position == "RB"),]
```

@@

```{r}
players %>%
  filter(position == "WR")

players %>%
  filter(position == "WR", age <= 26)

players %>%
  filter(position == "WR" | age >= 26)
```

@@

```{r}
players[c(3,4), c(2,4)]
players[subsetRows, subsetVars]
players[which(players$position == "RB"), subsetVars]
```

@@

```{r}
players %>%
  filter(position == "RB") %>%
  select(all_of(subsetVars))
```

@@

View(players)
```

@@

```{r}
players$newVar <- NA
```

@@

```{r}
players$newVar <- 1:nrow(players)
```

@@

players$newVar <- NULL
```

@@

```{r}
players$oldVar1 <- NA
players$oldVar1[which(players$position == "QB")] <- "quarterback"
players$oldVar1[which(players$position == "RB")] <- "running back"
players$oldVar1[which(players$position == "WR")] <- "wide receiver"
players$oldVar1[which(players$position == "TE")] <- "tight end"

players$oldVar2 <- NA
players$oldVar2[which(players$age < 30)] <- "young"
players$oldVar2[which(players$age >= 30)] <- "old"
```

@@

```{r}
players %>%
  mutate(across(c(
    oldVar1:oldVar2),
    ~ case_match(
      .,
      c("quarterback","old","running back") ~ 0,
      c("wide receiver","tight end","young") ~ 1)))
```

@@

```{r}
players <- players %>% 
  rename(
    newVar1 = oldVar1,
    newVar2 = oldVar2)
```

@@

players <- players %>% 
  rename_with(~ varNamesTo, all_of(varNamesFrom))
```

@@

```{r}
players$factorVar <- factor(players$ID)
players$numericVar <- as.numeric(players$age)
players$integerVar <- as.integer(players$newVar)
players$characterVar <- as.character(players$newVar1)
```

@@

```{r}
players %>%
  mutate(across(c(
    ID,
    age),
    as.numeric))

players %>%
  mutate(across(
    age:newVar1,
    as.character))

players %>%
  mutate(across(where(is.factor), as.character))
```

@@

Here are the data in the `players` object:

```{r}
players

dim(players)
```

The data are structured in ID form.
That is, every row in the dataset is uniquely identified by the variable, `ID`.

Here are the data in the `fantasyPoints` object:

```{r}
fantasyPoints

dim(fantasyPoints)
```

@@

```{r}
fullJoinData <- full_join(
  players,
  fantasyPoints,
  by = "ID")

fullJoinData

@@

```{r}
leftJoinData <- left_join(
  players,
  fantasyPoints,
  by = "ID")

leftJoinData

@@

```{r}
rightJoinData <- right_join(
  players,
  fantasyPoints,
  by = "ID")

rightJoinData

@@

```{r}
innerJoinData <- inner_join(
  players,
  fantasyPoints,
  by = "ID")

innerJoinData

@@

```{r}
semiJoinData <- semi_join(
  players,
  fantasyPoints,
  by = "ID")

semiJoinData

@@

```{r}
antiJoinData <- anti_join(
  players,
  fantasyPoints,
  by = "ID")

antiJoinData

@@

```{r}
crossJoinData <- cross_join(
  players,
  fantasyPoints)

crossJoinData

dim(crossJoinData)
```

@@

Here are the data in the `nfl_actualStats_offense_weekly` object.
The data are structured in player-season-week form.
That is, every row in the dataset is uniquely identified by the variables, `player_id`, `season`, and `week`.

@@

```{r}
dataLong <- nfl_actualStats_offense_weekly %>% 
  select(player_id, player_display_name, season, week, fantasy_points)

@@

Below, we widen the data widened by two variables (`season` and `week`), using `tidyverse`, so that the data are now in player form (where each row is uniquely identified by the `player_id` variable):

```{r}
dataWide <- dataLong %>% 
  pivot_wider(
    names_from = c(season, week),
    names_glue = "{.value}_{season}_week{week}",
    values_from = fantasy_points)

@@

```{r}
dataWide <- nfl_actualStats_offense_weekly %>% 
  select(player_id, player_display_name, season, week, recent_team, opponent_team)

@@

```{r}
dataLong <- dataWide %>% 
  pivot_longer(
    cols = c(recent_team, opponent_team),
    names_to = "role",
    values_to = "team")

@@

Specify scoring settings:
